# Replication

The replication model in `R/functions.R` is:

```r
fixest::feols(
  share_female ~ i(year, treatment_ban, ref = 2021) | institution + year,
  cluster = ~ institution,
  data = panel
)
```

This estimates year-specific differences between ban and control states, with 2021 as the reference year. Institution fixed effects absorb time-invariant differences across schools. Year fixed effects absorb shocks common to all schools.

## Run the template

```{r}
#| eval: false
source("R/run_replication_template.R")
```

Then inspect:

```{r}
#| eval: false
readr::read_csv("results/event_study_toy.csv")
```

## What counts as a successful replication?

A successful replication is not just "my coefficient matches." It includes:

- A complete row audit showing the schools and years in the sample.
- A treatment-status audit by state.
- A plot with 2021 normalized to zero.
- A table with estimates, standard errors, confidence intervals, and p-values.
- A written comparison to the Walker result.

## Interpretation template

Use this structure:

> Relative to 2021, the estimated 2022 difference between ban-state and control-state schools is `X`. This means the female applicant share was `X * 100` percentage points lower/higher in ban states than the control trend would predict. The pre-2022 coefficients are useful because they show whether the two groups were moving similarly before the policy shock.

