# Walker

This chapter is the first replication unit. Students replicate Walker et al. [-@walker2023anticipatory], then extend the design by adding Title IX pregnancy and parenting policy information from The Pregnant Scholar.

## Why This Paper

Walker et al. study a question that is concrete, recent, and methodologically useful: did the anticipated repeal of *Roe v. Wade* affect where female students applied to college?

The paper is short enough for students to read closely, but it contains the full empirical workflow:

- a defined population of schools
- a public data source
- a treatment and control comparison
- a panel dataset
- fixed effects
- an event-study figure
- robustness and subgroup analyses

That makes it a good first replication because students can see the whole machine at once.

## The Research Question

The core question is:

> After *Dobbs* became reasonably anticipated, did the share of female applicants fall at selective universities in states where abortion would be banned or restricted, relative to schools in states where abortion remained legal or bans were blocked?

The outcome is the share of first-time, first-year applicants who are female:

```text
female applicants / (female applicants + male applicants)
```

The main comparison is between schools in ban states and schools in control states.

## What The Paper Finds

Walker et al. report that in 2022, the share of female freshman applicants fell by about 0.9 percentage points in ban states relative to control states. Their estimate is approximately `-0.009`.

The paper interprets this as evidence that undergraduate applicants were sensitive to state reproductive health policy when deciding where to apply.

In this chapter, students should not treat the estimate as a number to copy. They should treat it as a claim to rebuild.

## The Replication Task

Students will create a school-year panel for 2018 through 2022 using Common Data Set admissions information.

The required raw data file is:

```text
data-raw/walker_cds_applicants.csv
```

It should contain:

```text
institution,state,year,female_applicants,male_applicants,us_news_rank,pct_out_state,source_url
```

The repository includes `data/example_cds_applicants.csv` so the R scripts run before the real data collection is complete. That file is toy data. It is for learning the workflow, not for making claims.

## Source Work

For each institution-year, students should collect:

- the school name
- the state
- the CDS year
- female applicants
- male applicants
- the source URL or local PDF path

Students should keep raw source files and notes separate from cleaned analysis files. If AI helps extract a number from a PDF, students still need to check the row against the original source.

## Treatment Definition

The replication uses a treatment indicator for states where abortion was banned or restricted after *Dobbs*, following the Walker paper's grouping.

The included file is:

```text
data/abortion_policy_status_2023.csv
```

This file gives each state a Walker-style status and a binary `treatment_ban` indicator.

## R Workflow

Run the replication template from the project root:

```{r}
#| eval: false
source("R/run_replication_template.R")
```

The script:

1. reads the real CDS file if it exists;
2. otherwise reads the toy CDS file;
3. merges the abortion policy status file;
4. calculates the female applicant share;
5. estimates the event-study model;
6. writes a results table and figure.

The main analysis model is:

```{r}
#| eval: false
fixest::feols(
  share_female ~ fixest::i(year, treatment_ban, ref = 2021) | institution + year,
  cluster = ~ institution,
  data = panel
)
```

This model compares ban-state and control-state schools in each year, using 2021 as the reference year. Institution fixed effects absorb stable differences across schools. Year fixed effects absorb shocks shared across schools.

## What Students Should Check

Before interpreting results, students should check:

- Do all schools have five years of data?
- Are any states missing treatment status?
- Did any join duplicate school-year rows?
- Is 2021 the omitted year?
- Are standard errors clustered by institution?
- Are the pre-2022 estimates close enough to support the parallel-trends story?

These checks matter more than a polished graph.

## Title IX Extension

After the replication, students extend the analysis with pregnancy and parenting policy information.

The Pregnant Scholar describes Title IX as a federal sex-discrimination law and provides resources for pregnant and parenting students [@pregnantscholar_basics]. Its policy directory lists campus policies such as pregnancy accommodations, breastfeeding accommodations, and parental leave [@pregnantscholar_policy].

Run:

```{r}
#| eval: false
source("R/scrape_pregnant_scholar_policies.R")
source("R/merge_title_ix_extension.R")
```

The scrape script attempts to read the live policy directory. If the page cannot be scraped, it uses the included teaching sample:

```text
data/title_ix_policy_directory_sample.csv
```

The merge script creates:

```text
data/analysis_title_ix_extension.csv
```

Students should interpret this source carefully. Being listed in The Pregnant Scholar policy directory is not the same thing as proving that every unlisted school lacks a policy. The measure should be described as directory listing status unless students complete a fuller policy audit.

## AI Use

AI is allowed in this chapter, but students must log it.

Good uses include:

- explaining the model formula
- drafting R checks for missing rows
- debugging a failed join
- extracting a draft row from a CDS PDF
- suggesting robustness checks

Bad uses include:

- inventing missing applicant counts
- citing a source that was not opened
- copying an interpretation without checking the coefficient
- letting AI decide the treatment definition

Students should record prompts in:

```text
prompts/ai_prompt_log_template.md
```

## Deliverables

By the end of the Walker chapter, students should submit:

- cleaned applicant panel
- event-study table
- event-study figure
- short replication memo
- Title IX extension file or table
- AI prompt log

The memo should answer one basic question: did their replication recover the direction, size, and logic of the Walker estimate?

