# Title IX Extension

The extension asks whether institutional pregnancy and parenting policy information changes the story.

The Pregnant Scholar explains Title IX as a federal sex-discrimination law and describes pregnancy, childbirth, lactation, miscarriage, abortion, and related conditions as covered pregnancy-related conditions [@pregnantscholar_basics]. Its policy directory lists campus policies such as parental leave, breastfeeding accommodations, and pregnancy accommodations [@pregnantscholar_policy].

## Research questions

Possible extension questions:

- Are schools with listed pregnancy accommodation policies less affected?
- Are listed policies more common in control states?
- Do ban-state schools with listed family-responsive policies look different from ban-state schools without listed policies?
- If using a literal religious-exemption dataset, are exempt institutions distributed differently across abortion policy environments?

## Scrape or snapshot

Run:

```{r}
#| eval: false
source("R/scrape_pregnant_scholar_policies.R")
```

If the page structure changes or no table is available, the script falls back to `data/title_ix_policy_directory_sample.csv`. That sample is only a teaching aid.

## Merge the extension

Run:

```{r}
#| eval: false
source("R/merge_title_ix_extension.R")
```

This creates `data/analysis_title_ix_extension.csv`.

## If "exemptions" means OCR religious exemptions

Use `data/religious_exemptions_template.csv`, collect the real source information, and save it as:

```text
data-raw/ocr_title_ix_religious_exemptions.csv
```

The merge script will add a `literal_title_ix_religious_exemption` flag when that file exists.

## Source limitations

The Pregnant Scholar policy directory is not a census of every college policy. It is a public directory and, on the current page, the project notes the directory and submission form are under review because of spam attacks. Treat directory presence as "listed in this source," not as "the school has a policy and every unlisted school does not."

